### 2023-03-09 paper.tex
write about benchmarking

While the primary purpose of VMRs is to provide better input for the distance calculations, their small size also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
Large 100~kb tiles will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.
To enable a more targeted characterization of 
The same principles may also be applied to DMRs, thus enabling in-depth characterization of epigenomic and gene regulatory differences between tissues, cell types or conditions.

@@

To demonstrate the value of our proposed improvements, we obtained two previously published scBS data sets and re-analyzed them.

\phantomsection
\subsection{Correlation of VMR methylation with gene expression}

The first data set used to test our methods is a single-cell multi-omics data set generated by \citet{kremer_scnmt}.
This data set comprises the methylomes of 1566 cells isolated from mouse forebrains, as well as single-cell transcriptomes of the same cells.

The first data set used for benchmarking is a single-cell multi-omics data set generated by \citet{argelaguet2019gastru}.
In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5--E7.5) as well as single-cell transcriptomes from the same cells.
In one of their analyses, the authors calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development.
The authors identified several genes where promoter methylation predicts gene expression, but obtained low correlations for the vast majority of genes.
We reasoned that VMRs obtained with the previously described sliding window approach might be a better predictor of gene expression.

In Figure \ref{fig:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with our approach.
This analysis revealed a number of genes whose expression does not correlate with promoter methylation, but correlates with VMR methylation.
This effect is also observed when using our proposed measure of methylation, the shrunken mean of residuals (Figure \ref{fig:correlation}B).

@@

To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{fig:correlation}A-B).
Dividing all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression (Figure \ref{fig:correlation}C).

\subsection{Identification of cellular sub-types}

\begin{figure}
    \begin{center}
        \includegraphics[width=\columnwidth]{figures/Fig_benchmark.png}
    \end{center}
    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.
    \textbf{(A)} Mean neighbor score obtained when analyzing data from \citet{kremer_scnmt} with different methods.
    Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA, labeled "pre-processing: ours") or using conventional methods (PCA on lightly imputed methylation fractions, based on \citep{luo2017single}).
    The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(B)}.
    \textbf{(C, D)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (C) or conventional methods based on genome tiling (D).}
    \label{fig:score}
\end{figure}

To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}.
This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or the shrunken means of residuals as a measure of methylation.

For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 regions with the highest sequencing coverage that were selected for PCA.

To assess performance of each combination of methods, we used a nearest neighbor count score (see Methods) that quantifies how well the different neuronal subtypes are separated in 15 dimensional PC space.
We used the original assignment of cells to cell types from \citet{luo2017single}.
The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{fig:score}A).
Most of this gain in performance can be attributed to the use of VMRs over genomic tiles.
The use of the shrunken means of residuals over methylation fractions provides an additional small improvement.

Since scBS protocols are still rather costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
Here, our approach keeps its good score, while the simple approach falls behind (Figure \ref{fig:score}B).

Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
In contrast, the neighbor count score for the top 1000 genomic 100~kb tiles is much lower.
To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{fig:score}C).
Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{fig:score}D).

>>>>

While the primary purpose of VMRs is to provide better input for the distance calculations, their small size also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
Large 100~kb tiles will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.
To enable a targeted characterization of epigenomic and gene regulatory differences between conditions or cell types, the same principles may also be applied to DMRs.

@@

To demonstrate the value of our proposed improvements, we obtained two previously published scBS data sets and re-analyzed them using our methods.

\phantomsection
\subsection{Correlating VMR methylation with gene expression}

The first data set used to test our methods is a single-cell multi-omics data set generated by \citet{kremer_scnmt}.
This data set comprises the methylomes of 1566 cells isolated from mouse forebrains, as well as matched single-cell transcriptomes of the same cells.
Among these cells are distinct cell types such as oligodendrocytes, oligodendrocyte precursor cells (OPCs) and endothelial cells, as well as cellular sub-states that are part of the continuous neural stem cell differentiation trajectory.
To assess whether our VMR detection method captures genomic intervals that are biologically meaningful, we probed whether their methylation level correlates with the expression of nearby genes.
Compared to promoter methylation, the obtained correlation coefficients are more extreme, indicating that VMR methylation often is a better predictor of gene expression than promoter methylation (Fig.~\ref{fig:correlation}A).
Indeed, a gene-wise comparison revealed many genes whose expression is correlated with the nearest VMR, but not with promoter methylation (Fig.~\ref{fig:correlation}B).
One such example gene, Htra1, is depicted in Fig.~\ref{fig:correlation}C.
While the promoter of this gene is lowly methylated regardless of gene expression, a VMR located downstream of the promoter is lowly methylated in cells with high Htra1 expression.

@@

\subsection{Effects of our methods on cell type identification}

\begin{figure}
    \begin{center}
        \includegraphics[width=\columnwidth]{figures/Fig_benchmark.png}
    \end{center}
    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.
    \textbf{(A)} Mean neighbor score obtained when analyzing data from \citet{kremer_scnmt} with different combinations of methods.
    Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA, labeled "pre-processing: ours") or using conventional methods (PCA on lightly imputed methylation fractions, based on \citep{luo2017single}).
    The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(B)}.
    \textbf{(C, D)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (C) or conventional methods based on genome tiling (D).}
    \label{fig:score}
\end{figure}

To assess the impact of our proposed methods, we next tested whether our methods improve the ability to distinguish cell types and cell states (Fig.~\ref{fig:score}).
To this end, we obtained cluster labels based on the single-cell transcriptomes from \citet{kremer_scnmt} (Fig.~\ref{fig:score}B).
We consider these cell labels as ground truth and next tested whether we are able to distinguish the same groups of cells based on their methylomes.
To do this, we subjected the methylomes to the standard scBS workflow:
defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
To determine genomic features for distance calculation, we either tiled the genome into 100~kb bins as is commonly done, or we used our VMR detection approach.
To quantify methylation in these intervals, we either averaged in these intervals, obtaining methylation percentages ("classic" pre-processing), or we calculated the shrunken mean of residuals.
Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}C,D).
While all cells form a continuous point cloud when using default methods, our improvements led to a clear separation of oligodendrocytes, OPCs and endothelial cells.
Furthermore, even cellular sub-types of cells in the continuous neural stem cell lineage were partially separated.

To quantify this performance gain in a more rigorious manner, we used a score that quantifies whether cells were placed in a neighborhood consisting of cells with the same cell type in 15-dimensional PC space ("neighbor score", see \ref{methods:score}).
Since scBS protocols are costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
To simulate smaller data sets, we thus sub-sampled the 1566-cell data set into smaller data sets, analyzed them with different method combinations, and calculated the mean neighbor score for each (Figure \ref{fig:score}A).
The results demonstrate that the forebrain cell clusters are more cleanly separated when using our proposed methods.
This performance gain was observed also in smaller data sets, although cell cluster separation generally becomes more difficult in those.
Both the use of VMRs instead of genomic tiles, as well as the use of the mean shrunken residuals over methylation averages improved cell cluster separation.
